---
title: "Fantastic Grids and Where to Find Them: Introduction to the octgridtools Package"
author: "S. Scott Whitmore"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fantastic Grids and Where to Find Them: Introduction to the octgridtools Package}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(octgridtools)
library(heyexr)
library(tidyverse)
library(scales)
```

`octgridtools` provides functions to (a) generate grids and (b) summarize 
spatial data using those grids. I developed the package to work specifically
with data from OCT segmentation, but most of the functions should be applicable
to any situation where you may want to analyze data using a grid.

## Creating grids

Currently, we have functions for working with three types of grids: sectoral,
rectangular, and hexagonal. Sectoral grids are currently specified by a TSV
file.

```{r createExamples}
# Sectoral grid
grid_file <- system.file(
  "extdata", "Grid_Circle_Macula_ETDRS.txt",
  package = "octgridtools"
)

grid_sectors <-
  generate_grid_sectors(
    read_tsv(grid_file),
    center_x = 0,
    center_y = 0
  ) %>%
  mutate(type = "sectors")
```

Rectangular and hexagonal grids can be constructed directly:

```{r createRectangles}
grid_rectangles <-
  generate_grid_rectangles(
    center_x = 0,
    center_y = 0,
    nrow = 11,
    ncol = 11,
    width = 6 / 11
  ) %>%
  mutate(type = "rectangles")

grid_hexagons <-
  generate_grid_hexagons(
    nrow = 11,
    ncol = 10,
    radius = 6 / 11
  ) %>%
  mutate(type = "hexagons") %>%
  # TASK: Move this into generate_grid_hexagons
  mutate(
    x = scales::rescale(x, to = c(-3, 3)),
    y = scales::rescale(y, to = c(-3, 3))
  )
```


Grids are stored as tibbles. Each grid contains one or more cells, and each cell
is made up of three or more nodes, defining a polygon. All grids must contain
the following columns:

* `.cell_id`
* `.node_id`
* `x`
* `y`

Other columns are optional.

* `row_id`
* `col_id`

## Plotting grids

Standard column makes it easy to combine multiple grids:

```{r figHorizontal, fig.width=7}
grid_examples <- bind_rows(grid_sectors, grid_rectangles, grid_hexagons)

p_grids <- 
  grid_examples %>%
  ggplot(aes(x = x, y = y)) +
  geom_polygon(
    aes(group = paste(type, .cell_id), 
        color = type),
    fill = NA
  ) +
  coord_equal()

p_grids + facet_grid(~type)
```

```{r figOverlay, fig.width=7}
p_grids
```

When generating plots, it can be helpful to label each cell. To calculate the 
center of each cell, compute the mean `x` and `y` value for each cell.

```{r figCenters, fig.width = 8, fig.height = 4}
centers <- 
  grid_examples %>%
  group_by(type, .cell_id) %>%
  dplyr::summarize(x = mean(x), y = mean(y)) %>%
  ungroup()

grid_examples %>%
  ggplot(aes(x = x, y = y)) +
  geom_polygon(aes(group = paste(type, .cell_id)),
               color = "black", 
               fill = NA) +
  geom_text(data = centers,
            mapping = aes(x = x, 
                          y = y, 
                          label = .cell_id, 
                          color = as.factor(.cell_id)),
            fontface = "bold",
            size = 2.5) +
  scale_color_discrete(guide = "none") +
  coord_equal() +
  facet_grid(~type)

# TASK: Ensure consistent number ordering for the rectangular and hexagonal
#       grids.
```


## Keeping track of scaling and coordinate spaces

TASK: Write up vignette for heyexr detailing the 0-based (OCTExplorer) vs. 
1-based system (R matrices, tables), plotting from bottom left vs. top left 
(via `scale_y_reverse()`). Summarize that information here.

* DISCUSS VOXEL SPACE VS ANATOMIC (METRIC) SPACE VS. SLO SPACE (MAY REQUIRE 
ROTATION OF VOXEL SPACE BSCAN COORDINATES.)

## Transforming grids

Grid can be transformed using (homogeneous) affine matrices or using the
`scales::rescale` function to scale or move the bounding box of the grid.

## Using grids to summarize or transform spatially organized data

1. Find overlaps between the coordinates of the grid cells and the data.
2. Group the data by their cells.
3. Perform summary statistics on the grouped data.
4. Merge the summary statistics with the original grid coordinates.
5. Display the grid, mapping summaries to grid polygon aesthetics.

In stead of summary statistics, you could perform other kinds of transforms on
the data within a cell region, and re-dispay the data. 

## Important points to consider when using grids to summarize data

* When analyzing images of paired organs, such as eyes, do you need to compare
comparable regions across the pair? For example, do you need to consider to
compare the temporal quadrant of the right (OD) eye to the temporal quadrant of 
the left eye? If so, then you'll need to reflect the grid prior so that the 
appropriate `.cell_id`s match anatomically. 
* Is the size of your grid cells smaller than the resolution of your data?
* Does the orientation of your grid match the orientation of your data? How is
your image or map oriented?




