---
title: "Understanding ETDRS Grid Analysis from OCTExplorer v 3.8.0 (9/8/2017)"
author: "S. Scott Whitmore"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options: 
  chunk_output_type: console
---

This vignette serves as the standard reference for comparing thickness analyses
computed with OCTExplorer vs. thickness analyses computed with `octgridtools`
functions. The data used to generate this package is contained in the `octdata`
package.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>",
  out.width = "100%")
knitr::opts_knit$set(root.dir = here::here())

setwd(here::here())

library(heyexr)
library(octgridtools)
library(tidyverse)
library(patchwork)
library(broom)

data(layer_info, package = "heyexr")

data(thickness_analysis, package = "octdata")

data(sdoct, package = "octdata")

example_info <- 
  octdata::sdoct_files %>%
  filter(subject_id == "TEST_T") %>% 
  filter(sample_id %in% names(sdoct)) %>% 
  inner_join(octdata::vol_headers) %>%
  filter(num_bscans == 61) %>%
  # Add in the screenshots from OCTExplorer
  inner_join(octdata::sdoct_files %>% 
               filter(file_type == "projected_image_custom") %>%
               select(sample_id, file_name) %>%
               rename(nfl_image = file_name))

example <- sdoct[example_info$sample_id]

names(example) <- example_info$scan_position

rm(sdoct)

aspect_ratio <- 
  example %>%
  map(~get_enface_aspect(.x$volume))

nfl_images <- 
  example_info$nfl_image %>%
  set_names(example_info$scan_position) %>%
  map(~magick::image_read(system.file("extdata", .x, package = "octdata")))
```


```{r figOctexplorerProjections, fig.cap="Projected images rendered by OCTExplorer of the nerve fiber layer (NFL; surfaces 1-2) in the ETDRS regions for right and left eyes.", fig.width=12, fig.height=6}
screenshot <-
  nfl_images %>%
  imap(
    ~tibble(
      x = c(1 - 0.5, example[[.y]]$volume$header$size_x + 0.5), 
      y = c(1 - 0.5, example[[.y]]$volume$header$num_bscans + 0.5)
      ) %>%
     ggplot(aes(x = x, y = y)) +
     geom_point(alpha = 0) +
     annotation_raster(
       raster = .x, 
       xmin = 1 - 0.5, 
       xmax = example[[.y]]$volume$header$size_x + 0.5,
       ymin = 1 - 0.5,
       ymax = example[[.y]]$volume$header$num_bscans + 0.5) +
     scale_x_continuous(expand = c(0, 0)) +
     scale_y_continuous(expand = c(0, 0)) +
      coord_fixed(ratio = aspect_ratio[[.y]])
      )

screenshot$OD + screenshot$OS + plot_layout(ncol = 2)
```

Which ETDRS grid IDs match which anatomic regions?

```{r results="asis"}
thickness_analysis %>% 
  inner_join(layer_info) %>%
  select(sample_id, laterality, oct_center_type, layer, region, mean, sd) %>%
  filter(layer == "NFL") %>%
  knitr::kable()
```

NOTE: Both images are centered on the macula, but for some reason the data
was not entered into Heidelberg Explorer correctly.m,,                                                           

This figure and table indicates the region 7 is located superior to the center
and region 9 is located inferior to the center. 

Let's read in the ETDRS grid definition I compiled a couple of years ago. How
does this natively appear when plotted with `ggplot`?

```{r figEtdrsGridOd, fig.caption="Nine grid regions for the Early Treatment of Diabetic Retinopathy Study (ETDRS) grid as used by OCTExplorer. Orientation for the right (OD) eye shown.", fig.width=12, fig.height=6}
# Sectoral grid
data(grid_sectors_etdrs, package = "heyexr") 

grid_sectors_etdrs <- 
  grid_sectors_etdrs %>%
  mutate(type = "sectors")

flip_bit <- -1

affine_flip_y <- matrix(c(flip_bit, 0, 0, 0, 1, 0, 0, 0, 1), 
                        ncol = 3, 
                        byrow = TRUE)

grid_sectors_etdrs_ou <- list() 
grid_sectors_etdrs_ou$OD <- grid_sectors_etdrs
grid_sectors_etdrs_ou$OS <- affine_transform_coord(grid_sectors_etdrs, 
                                                   affine = affine_flip_y)

grid_sectors_etdrs_ou_tbl <-
  grid_sectors_etdrs_ou %>%
  map_dfr(~.x, .id = "laterality")

grid_sectors_etdrs_centers <-
  grid_sectors_etdrs_ou_tbl %>%
  group_by(laterality, .cell_id) %>%
  dplyr::summarise(x = mean(x), y = mean(y))

grid_sectors_etdrs_ou_tbl %>%
  ggplot(aes(x = x, y = y)) +
  geom_polygon(aes(group = .cell_id),
               color = "black",
               fill = NA) +
  geom_text(data = grid_sectors_etdrs_centers,
            mapping = aes(x = x, y = y, label = .cell_id)) +
  coord_equal() +
  labs(title = "ETDRS\ndefault y-axis") +
  facet_grid(~ laterality)
```

When plotting using ggplot, 7 appears superior to the fovea and 9 appears 
inferior to the fovea, as in the figure above. 

Things to remember:

1. The ETDRS grid definition assumes OD (right) orientation. The grid must be
mirrored about the superior-inferior axis to be used with OS (left) orientation.

2. The grid assumes that positive values on the y-axis are superior to the 
center and negative values on the y-axis are inferior to the center.

Overlay the grid on the structural en face image.

```{r figStructuralEnface, fig.cap="Structural en face images of the retina and RPE (surfaces 1 - 10).", fig.width=12, fig.height=6}
seg_volume <- 
  example %>%
  map(
    ~expand_surfaces_to_volume(
      surface_array = iowa_segmentation_to_array(.x$segmentation),
      vol_dim = dim(.x$volume$bscan_images)
    )
  )

surface_array <- 
  example %>%
  map(~iowa_segmentation_to_array(.x$segmentation))

slicer_full_thickness <- 
  seg_volume %>%
  map(~create_slicer_range(seg_volume = .x, 1, 10))

volume_sliced <- 
  example %>%
  map2(slicer_full_thickness, ~.x$volume$bscan_images * .y)

enface_full_thickness <- 
  volume_sliced %>%
  map2(
    example,
    # Using gamma correction here as I expect that is what OCTExplorer uses.
    ~gamma_correction_array(.x) %>%
      apply(c(1,2), sum, na.rm = TRUE) %>%
      melt_array(c("x", "y", "value")) %>%
      add_undefined(get_undefined(.y$segmentation)))

p_enface_full_thickness <- 
  enface_full_thickness %>%
  map(~.x %>% 
    ggplot(aes(x = x, y = y)) +
    geom_raster(aes(fill = value_defined)) +
    scale_fill_gradient(guide = "none", low = "black", high = "white") +
    scale_x_continuous(expand = c(0, 0)) + 
    scale_y_continuous(expand = c(0, 0)) +
    coord_fixed(ratio = aspect_ratio$OD) 
  )
  
p_enface_full_thickness$OD + 
  p_enface_full_thickness$OS + 
  plot_layout(ncol = 2)
```

```{r figStructuralEnfaceGrid, fig.cap="Grid regions shown overlaid on right and left eyes.", fig.width=12, fig.height=6}
# Now convert the grid to a slicer
grid_slicer <- 
  example %>% 
    map2(grid_sectors_etdrs_ou,
    ~convert_grid_to_slicer(
    grid = .y, 
    grid_center = .x$grid_center, 
    dims = dim(.x$volume$bscan_images), 
    scale_x = .x$volume$header$scale_x,
    scale_y = .x$volume$header$distance
    )
  )

p_overlay <- 
  grid_slicer %>%
  map(
    ~.x[ , , 1] %>% 
    melt_array(c("x", "y", ".cell_id")) %>%
    
    ggplot(aes(x = x, y = y)) + 
    geom_raster(aes(fill = as.factor(.cell_id)),
                alpha = 0.7) +   
    scale_x_continuous(expand = c(0,0)) + 
    scale_y_continuous(expand = c(0,0)) +
    scale_fill_brewer(name = "region", palette = "Set1"),
    .id = laterality
    )

p_overlay$OD + 
  p_overlay$OS + 
  plot_layout(ncol = 2) & 
  coord_fixed(ratio = aspect_ratio$OD)

annotation_add_enface <- function(plot_overlay, width, height) {
  temp_png <- tempfile()
  
  dpi <- 72
  
  ggsave(plot_overlay, 
         filename = temp_png, 
         device = "png", 
         units = "in",
         width = width * 2 / dpi,
         height = height * 2 / dpi, 
         bg = "transparent") 
  
  overlay <- magick::image_read(temp_png)
  
  unlink(temp_png)
  
  annotation_raster(raster = overlay, 
                    xmin = 1,
                    xmax = width,
                    ymin = 1,
                    ymax = height, 
                    interpolate = FALSE)
}

p_enface_full_thickness_overlay <-
  p_enface_full_thickness  %>%
  imap(~.x + 
    annotation_add_enface(p_overlay[[.y]] + theme_nude(), 
                          width = example[[.y]]$volume$header$size_x,
                          height = example[[.y]]$volume$header$num_bscans)
  )

p_enface_full_thickness_overlay$OD + 
  p_enface_full_thickness_overlay$OS + 
  plot_layout(ncol = 2)
```

Notice that the en face image computed by OCTExplorer displays greater contrast
than the en face image I'm computing here. 

Now let's compute the ETDRS thickness using this grid defintion for the example
eye and plot the results:

```{r results="asis"}
grid_thickness <- 
  example %>%
  map(
    ~calculate_grid_thickness(
      segmentation = .x$segmentation, 
      grid_center = .x$grid_center, 
      grid = grid_sectors_etdrs, 
      flip_region = "OD"
      ) %>%
  left_join(layer_info) 
  ) 

grid_thickness %>%
  bind_rows(.id = "laterality") %>%
  filter(layer == "NFL") %>%
  select(laterality, layer, .cell_id, thickness_um_mean, thickness_um_sd) %>%
  mutate(thickness_um_mean = round(thickness_um_mean, digits = 2),
         thickness_um_sd = round(thickness_um_sd, digits = 2)) %>%  
  knitr::kable()
```

NOTE: The slicer concept assumes the a voxel can be assigned to one and only one
"level" in the slicer "factor".

Let's overlay this information back on the en face graph.

```{r figStructuralEnfaceGridThickness, fig.cap="ETDRS grid regions shown overlaid on right and left eyes. Within each region, the region number, mean thickness, and standard deviation of thickness are shown.", fig.width=12, fig.height=12}
grid_thickness_enface_centers <- 
  grid_slicer %>% 
    map2(grid_thickness,
         ~.x[ , , 1] %>% 
    melt_array(c("x", "y", ".cell_id")) %>%
    group_by(.cell_id) %>%
    dplyr::mutate(voxel_center_x = mean(x), 
                  voxel_center_y = mean(y)) %>%
    ungroup() %>%
    left_join(.y)
  )

# Approximate the OCTExplorer color palette using these colors.
# TASK: See if these colors are defined in OCTExplorer
custom_colors <- c("blue", "lightblue", "green", "yellow", "orange", "red")

p_overlay_2 <- 
  grid_thickness_enface_centers %>%
  map(~.x %>%
  filter(layer == "NFL" | is.na(.cell_id)) %>%
  
  ggplot(aes(x = x, y = y)) + 
  geom_raster(aes(fill = thickness_um_mean),
              alpha = .75) +   
  scale_x_continuous(expand = c(0,0)) + 
  scale_y_continuous(expand = c(0,0)) +
  scale_fill_gradientn(name = "thickness (µm)",
                       colors = custom_colors, na.value = NA)
  )

p_enface_full_thickness_overlay_2 <-
  p_enface_full_thickness %>%
  imap(~.x +
    annotation_add_enface(p_overlay_2[[.y]] + theme_nude(), 
                          width = example[[.y]]$volume$header$size_x,
                          height = example[[.y]]$volume$header$num_bscans) +    
    geom_point(x = example[[.y]]$grid_center$center$x,
               y = example[[.y]]$grid_center$center$z,
               color = "red",
               size = 2) +
    # geom_vline(xintercept = example[[.y]]$grid_center$center$x,
    #            color = "green") +  
    # geom_hline(yintercept = example[[.y]]$grid_center$center$z,
    #            color = "red") +
    geom_text(data = grid_thickness_enface_centers[[.y]] %>%
                filter(layer == "NFL") %>%
                select(.cell_id, 
                       voxel_center_x, 
                       voxel_center_y, 
                       thickness_um_mean, 
                       thickness_um_sd) %>%
                distinct() %>%
                mutate(thickness_um_mean = round(thickness_um_mean, digits = 2),
                       thickness_um_sd = round(thickness_um_sd, digits = 2)),
              
              mapping = aes(x = voxel_center_x, 
                            y = voxel_center_y, 
                            label = paste0(.cell_id, "\n", 
                                          thickness_um_mean, 
                                          "\n(",
                                          thickness_um_sd,
                                          ")")),
              fontface = "bold",
              lineheight = 0.75) +
      ggtitle(paste0("(", example[[.y]]$grid_center$center$x, ", ", 
                     example[[.y]]$grid_center$center$z, ")"))
  )

screenshot$OD + 
  screenshot$OS + 
  p_enface_full_thickness_overlay_2$OD +
  p_enface_full_thickness_overlay_2$OS +
  plot_layout(ncol = 2, nrow = 2, byrow = TRUE)
```

**Conclusion**: ~~The thickness values computed by OCTExplorer and the 
`octgridtools` function are very close. I suspect there are three sources of
differences between the two functions:~~ Update: Now that I've revised how I'm
reading grid center XML files, my results are stunningly close. Slight 
differences are likely rounding error.

~~1. Which voxels are assigned to which regions.~~ Update: The difference may also be due to a difference in the number of points 
I'm using to define a circle or an arc. Increasing point densities along the 
edge of a polygonal region will change which voxel is counted within or outside
the edge. Again, this is a hunch as to why there might be 0.01 difference between
some of the calculations.
2. When values are scaled (before or after computing the average).
3. How many significant digits are retained between steps.

```{r rSessionInfo}
sessionInfo()
```
